Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-10-26 Manuscript.Rmd
Spell check.

In many parts of Europe, measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides since the late 1980s, with an according decrease in N deposition [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased N deposition. Among others, N deposition is still high at many sites, since in contrast to nitrogen oxides, there was only a minor recent decrease of ammonia emissions. Thus, even if N deposition is reduced, large areas might still be above the critical threshold above which harmful effects on plant diversity do occur [i.e. above the critical load for N deposition; @Bobbink2010; @Slootweeg2015; @RihmAchermann2016]. Furthermore, there is a possibility that communities reach an alternative stable state after decades of increased N deposition and that the respective plant species are unlikely to disappear even if N deposition is reduced [@Stevens2016]. Or, if oligotrophic species desappeared from the entire landscape, dispersal limitation may prevent oligotrophic species from recolonizing sites [@Dirnbock2004]. Therefore, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of plant communities.

Recovery of existing plant communities after high N deposition would imply that the state of communities measured at different points in time is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness, a biodiversity endpoint that can be assessed and communicated relatively easily, is often negatively related to nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favorable changes can be derived from the traits of the species in a community. In Europe, environmental preference of plants has often been expressed using indicator values assigned to each plant species [@Ellenberg1992; @Landolt2010]. Examples of such metrics would be the number of oligotrophic species, or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus, a useful approach to understanding biodiversity change is through estimates of species turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction are compared to random expectation [@Chase2011].

@@

## Monitoring data and community measures {-}
We analyzed the presence/absence of vascular plants sampled in the Swiss Biodiversity Monitoring (BDM) program that was launched in 2001 to monitor Switzerland's biodiversity and to comply with the Convention on Biological Diversity of Rio de Janeiro [@Weber2004]. The sampling sites were circles with a size of 10 m$^2$, and data collection was carried out by qualified botanists who visited each sampling site twice within the same season. During each visit, all vascular plant species detected on the plot were recorded except for young plants that have not yet eveloped at least the first pair of leaves after the cotyledons. For details on the field methods see @Plattner2004, @Roth2013 and @Roth2017.

After the sampling of the plant data, the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analyzed the data from 2003 to 2017. During that study period, each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012, and the third from 2013 to 2017. These selection criteria yielded `r nrow(sites)` sites that had been sampled three times during 2003-2017. The distribution of sites accross Switzerland is given in Fig. \@ref(fig:studysite).

@@

Before the analyses we removed all records that were not identified on species level. For each survey (that consisted of two visits per season) we then calculated the following biodiversity endpoints [@Rowe2017]: We used the number of recorded species (species richness) that can be easily related to many conservation targets [@Rowe2017]. Additionally, we calculated the community mean of the Landolt indicator values of recorded species. Similar to the Ellenberg’s indicator values [@Ellenberg1974], the Landolt values are ordinal numbers that express the realized ecological optima of plants species for different climatic, soil or land-use variables. The Landolt indicator values were developed for the specific situation in Switzerland, published the first time in @Landolt1977 and recalibrarted in @Landolt2010. Their predictive power was tested in different studies [e.g. @Scherrer2011]. We analysed the indicator values for temperature (1: high elevation species; 5: low elevation species), soil moisture (1: species that grow in soils with low moisture; 5: species that grow in water-saturated soils), nutrients (referring in particular to nitrogen, but also to phosphorus; 1: species that grow in nutrient-poor soils; 5: species that grow under nutrient-rich conditions) and light (1: species that grow in shade; 5: species that predominantly occur in bright places).

@@

We used logistic GLMMs to analyze whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this change depended on the species indicator value for nutrients. For the analysis of the colonisation probability we examined for all species whether they were also observed in the subsequent survey (1st vs 2nd, 2nd vs 3rd). The variable ‘occ’ (observed = 1; not observed = 0) was then used as dependent variable in the logistic GLMM. As predictor variables we used the N deposition of the site, the average indicator value for nutrients, and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic GLMM was also used to investigate local survival probabilities. In that case, however, we selected all species that were recorded during the first or second survey, and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

@@

## Data accessibility and reproducibility of results {-}
Data and R Markdown documents (Manuscript.Rmd and Appendix_A.Rmd) to fully reproduce this manuscript including figures and tables are provided at https://github.com/TobiasRoth/NDep-Trend. An R Markdown document is written in markdown (plain text format) and contains chunks of embedded R code to produce the figures and tables [@Xie2018]. Raw data for analyses are provided in the folder 'RData' and the folder 'R' contains the R-Script that was used to export the data from the BDM database. The folder "Settings" contains a list of all the R packages (including version number) that were in the workspace when the manuscript was rendered. Note, however, that the files on the github repository might have slightly changed since submitting the mansuscript. The version of all files will be archived on Zenodo upon acceptance of the manuscript.

@@

## Plant communities {-}
In total, `r n_distinct(pl$aID_SP)` plant species were recorded on the `r nrow(sites)` plots. Including the data of all three visits, `r mean(surv$SR)` $\pm$ `r sd(surv$SR)` (average $\pm$ SD) species were observed per plot. The lowest number of species recorded during a survey was `r min(surv$SR)` species and the highest number was `r max(surv$SR)` species. In average `r mean(100 * surv$KL)` $\pm$ `r sd(100 * surv$KL)`% of the recorded species were annual species with one plot reaching up to `r max(100 * surv$KL)`% annual species. The average indicator value for temperature accross all surveys was `r mean(surv$T)` $\pm$ `r sd(surv$T)` ranging from `r min(surv$T)` to `r max(surv$T)`. The average indicator value for soil moisture was `r mean(surv$F)` $\pm$ `r sd(surv$F)` ranging from `r min(surv$F)` to `r max(surv$F)`. The average indicator value for nutrients was `r mean(surv$N)` $\pm$ `r sd(surv$N)` ranging from `r min(surv$N)` to `r max(surv$N)`. And the average indicator value for light was `r mean(surv$L)` $\pm$ `r sd(surv$L)` ranging from `r min(surv$L)` to `r max(surv$L)`.

@@

While we could not detect a consistent decrease in the average indicator values for nutrients (Table \@ref(tab:communitytrendstab)), the nutrient value of colonizing species was below average (i.e. a strongly negative effect size) at low deposition rate and the difference to random expectation became smaller along  the N deposition gradient (second last line in Table \@ref(tab:difffromrandomtab)). This hihger colonization rate of species with low nutrient value at sites with low N deposition rate might have affected the spatial pattern of oligotrophic species richness: sites with low N deposition were likely to become more species-rich over time. This likely resulted in a steeper slope of the negative relationship between N deposition and oligotrophic species richness, when comparing this relationship with a spatial approach as in @Roth2013. Indeed, if we apply such an approach at different points in time to infer the effects of N deposition on the spatial variation of oligotrophic species richness, the resulting effect size (i.e., the slope) became more negative over time (Fig. \@ref(fig:figconsequences)).

@@

In Swiss mountain hay meadows the average N deposition was still rather high with an average of `r mean(sites$NTOT_2015)` kg ha$^{-1}$ yr$^{-1}$), which is at the upper limit of the suggested critical load [@Roth2017]. Furthermore, Nitrogen deposition only weakly decreased by about `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ between 2000 and 2015. This is only about one tenth of the decrease in England, where N deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. The still comparatively high N deposition rate and the rather low decrease in N deposition combined with the fact that most of the species are perennials likely explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled effects of N depositon on community composition. In particular, we found that the species disappearing from the sites tended to have below average indicator values for temperature, while species that newly colonize sites had above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizing species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how N deposition is affecting disappearance and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effects of disappearing and colonizing species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at higher N deposition rates. This suggests that mountain hay meadow communities can reach alternative stable states, with eutrophic species that are unlikely to disappear even if N deposition is reduced [@Stevens2016]. Taken together, these reasons might explain why the composition of mountain hay meadow communities responded stronger to climate warming than to nitrogen reduction, although the reduction in nitrogen resulted in above average colonization of oligotrophic species.

Our results conform to the results of recent reviews on biodiversity change suggesting that local-scale species communities are often undergoing profound changes but not systematic loss [@Dornelas2014]. However, our comparision with replicated surveys from the same year warn that an important portion of the observed turnover of species might be due to pseudo-turnover (i.e. species difference between two surveys that were conducted by two surveyors during the same year on the same site). Given that the BDM program had invested significantly in developing reproducable methods and has continously invested in quality control [@Plattner2004], the recorded pseudo-turnover was quite high. A potential explanation is that the replicated surveys are not conducted during the same days and in few cases the situation might have changeed profoundly for example because the meadow was cut between the surveys. Furthermore, one of the botanist may not have identified difficult to identify species on species-level (i.e. differences in botanical skills), and--because we analysed only records that were identified on species level--it increased pseudo-turnover even tough both botanists detected the species. This must be taken into account when evaluating the presented results. For example, we find that sites with high species richness have higher turnover than sites with a low species richness. This seems biologically plausible, since the average species coverage in species-rich sites must be lower than in species-poor sites and species with low coverage probably have a higher turnover. At the same time, however, the result could also be simply due to pseudo-turnover, since species identification in species-rich sites is probably more difficult than in species-poor sites. However, for the presented results that suggest differences in colonization or local survival in relation to the species indicator values we can hardly imagine how this could be caused by methodological issues. We are thus confident that most of the results presented in this study are largely unbiased.

@@

However, the results of Fig. \@ref(fig:figconsequences) also show that the results of spatial comparisons must be interpreted carefully. When we applied inferred the spatial variation in oligotrhophic species richness with the same covariates as in Roth et al. (2017) for different study periods, to infer how oligotrophic species richness is decreasing along the N deposition gradient, the relationship appeared to vary between study periods. The decrease gradually became steeper (more negative) over time, except for the first two study periods. Our first interpretation was that the N deposition effect became stronger over time. This was against our prediction that the effect of N deposition should become weaker over time, since N deposition is decreasing during the study period. Then we realized that species turnover was highest at low N deposition sites (Tab. \@ref(tab:turnovertab)). At low N deposition rates, colonizing species have below average indicator values for nutrients. It seems that the decrease in N deposition resulted in oligotrophic species replacing eutrophic species particularly at sites with low N deposition. This seems to explain why the decline in oligotrophic species richness inferred from spatial patterns of species richness and N deposition is becoming steeper over time. And this may be interpreted as evidence that plant communities are recovering at least at low deposition sites, and that the negative N deposition effects have not become stronger over time.

>>>>

In many parts of Europe, measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides since the late 1980s, with an according decrease in N deposition [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased N deposition. Among others, N deposition is still high at many sites, since in contrast to nitrogen oxides, there was only a minor recent decrease of ammonia emissions. Thus, even if N deposition is reduced, large areas might still be above the critical threshold above which harmful effects on plant diversity do occur [i.e. above the critical load for N deposition; @Bobbink2010; @Slootweeg2015; @RihmAchermann2016]. Furthermore, there is a possibility that communities reach an alternative stable state after decades of increased N deposition and that the respective plant species are unlikely to disappear even if N deposition is reduced [@Stevens2016]. Or, if oligotrophic species disappeared from the entire landscape, dispersal limitation may prevent oligotrophic species from recolonizing sites [@Dirnbock2004]. Therefore, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of plant communities.

Recovery of existing plant communities after high N deposition would imply that the state of communities measured at different points in time is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness, a biodiversity endpoint that can be assessed and communicated relatively easily, is often negatively related to nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference of plants has often been expressed using indicator values assigned to each plant species [@Ellenberg1992; @Landolt2010]. Examples of such metrics would be the number of oligotrophic species, or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus, a useful approach to understanding biodiversity change is through estimates of species turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction are compared to random expectation [@Chase2011].

@@

## Monitoring data and community measures {-}
We analysed the presence/absence of vascular plants sampled in the Swiss Biodiversity Monitoring (BDM) program that was launched in 2001 to monitor Switzerland's biodiversity and to comply with the Convention on Biological Diversity of Rio de Janeiro [@Weber2004]. The sampling sites were circles with a size of 10 m$^2$, and data collection was carried out by qualified botanists who visited each sampling site twice within the same season. During each visit, all vascular plant species detected on the plot were recorded except for young plants that have not yet developed at least the first pair of leaves after the cotyledons. For details on the field methods see @Plattner2004, @Roth2013 and @Roth2017.

After the sampling of the plant data, the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analysed the data from 2003 to 2017. During that study period, each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012, and the third from 2013 to 2017. These selection criteria yielded `r nrow(sites)` sites that had been sampled three times during 2003-2017. The distribution of sites across Switzerland is given in Fig. \@ref(fig:studysite).

@@

Before the analyses we removed all records that were not identified on species level. For each survey (that consisted of two visits per season) we then calculated the following biodiversity endpoints [@Rowe2017]: We used the number of recorded species (species richness) that can be easily related to many conservation targets [@Rowe2017]. Additionally, we calculated the community mean of the Landolt indicator values of recorded species. Similar to the Ellenberg’s indicator values [@Ellenberg1974], the Landolt values are ordinal numbers that express the realized ecological optima of plants species for different climatic, soil or land-use variables. The Landolt indicator values were developed for the specific situation in Switzerland, published the first time in @Landolt1977 and recalibrated in @Landolt2010. Their predictive power was tested in different studies [e.g. @Scherrer2011]. We analysed the indicator values for temperature (1: high elevation species; 5: low elevation species), soil moisture (1: species that grow in soils with low moisture; 5: species that grow in water-saturated soils), nutrients (referring in particular to nitrogen, but also to phosphorus; 1: species that grow in nutrient-poor soils; 5: species that grow under nutrient-rich conditions) and light (1: species that grow in shade; 5: species that predominantly occur in bright places).

@@

We used logistic GLMMs to analyse whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this change depended on the species indicator value for nutrients. For the analysis of the colonisation probability we examined for all species whether they were also observed in the subsequent survey (1st vs 2nd, 2nd vs 3rd). The variable ‘occ’ (observed = 1; not observed = 0) was then used as dependent variable in the logistic GLMM. As predictor variables we used the N deposition of the site, the average indicator value for nutrients, and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic GLMM was also used to investigate local survival probabilities. In that case, however, we selected all species that were recorded during the first or second survey, and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

@@

## Data accessibility and reproducibility of results {-}
Data and R Markdown documents (Manuscript.Rmd and Appendix_A.Rmd) to fully reproduce this manuscript including figures and tables are provided at https://github.com/TobiasRoth/NDep-Trend. An R Markdown document is written in markdown (plain text format) and contains chunks of embedded R code to produce the figures and tables [@Xie2018]. Raw data for analyses are provided in the folder 'RData' and the folder 'R' contains the R-Script that was used to export the data from the BDM database. The folder "Settings" contains a list of all the R packages (including version number) that were in the workspace when the manuscript was rendered. Note, however, that the files on the github repository might have slightly changed since submitting the manuscript. The version of all files will be archived on Zenodo upon acceptance of the manuscript.

@@

## Plant communities {-}
In total, `r n_distinct(pl$aID_SP)` plant species were recorded on the `r nrow(sites)` plots. Including the data of all three visits, `r mean(surv$SR)` $\pm$ `r sd(surv$SR)` (average $\pm$ SD) species were observed per plot. The lowest number of species recorded during a survey was `r min(surv$SR)` species and the highest number was `r max(surv$SR)` species. In average `r mean(100 * surv$KL)` $\pm$ `r sd(100 * surv$KL)`% of the recorded species were annual species with one plot reaching up to `r max(100 * surv$KL)`% annual species. The average indicator value for temperature across all surveys was `r mean(surv$T)` $\pm$ `r sd(surv$T)` ranging from `r min(surv$T)` to `r max(surv$T)`. The average indicator value for soil moisture was `r mean(surv$F)` $\pm$ `r sd(surv$F)` ranging from `r min(surv$F)` to `r max(surv$F)`. The average indicator value for nutrients was `r mean(surv$N)` $\pm$ `r sd(surv$N)` ranging from `r min(surv$N)` to `r max(surv$N)`. And the average indicator value for light was `r mean(surv$L)` $\pm$ `r sd(surv$L)` ranging from `r min(surv$L)` to `r max(surv$L)`.

@@

While we could not detect a consistent decrease in the average indicator values for nutrients (Table \@ref(tab:communitytrendstab)), the nutrient value of colonizing species was below average (i.e. a strongly negative effect size) at low deposition rate and the difference to random expectation became smaller along  the N deposition gradient (second last line in Table \@ref(tab:difffromrandomtab)). This higher colonization rate of species with low nutrient value at sites with low N deposition rate might have affected the spatial pattern of oligotrophic species richness: sites with low N deposition were likely to become more species-rich over time. This likely resulted in a steeper slope of the negative relationship between N deposition and oligotrophic species richness, when comparing this relationship with a spatial approach as in @Roth2013. Indeed, if we apply such an approach at different points in time to infer the effects of N deposition on the spatial variation of oligotrophic species richness, the resulting effect size (i.e., the slope) became more negative over time (Fig. \@ref(fig:figconsequences)).

@@

In Swiss mountain hay meadows the average N deposition was still rather high with an average of `r mean(sites$NTOT_2015)` kg ha$^{-1}$ yr$^{-1}$), which is at the upper limit of the suggested critical load [@Roth2017]. Furthermore, Nitrogen deposition only weakly decreased by about `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ between 2000 and 2015. This is only about one tenth of the decrease in England, where N deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. The still comparatively high N deposition rate and the rather low decrease in N deposition combined with the fact that most of the species are perennials likely explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled effects of N deposition on community composition. In particular, we found that the species disappearing from the sites tended to have below average indicator values for temperature, while species that newly colonize sites had above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizing species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how N deposition is affecting disappearance and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effects of disappearing and colonizing species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at higher N deposition rates. This suggests that mountain hay meadow communities can reach alternative stable states, with eutrophic species that are unlikely to disappear even if N deposition is reduced [@Stevens2016]. Taken together, these reasons might explain why the composition of mountain hay meadow communities responded stronger to climate warming than to nitrogen reduction, although the reduction in nitrogen resulted in above average colonization of oligotrophic species.

Our results conform to the results of recent reviews on biodiversity change suggesting that local-scale species communities are often undergoing profound changes but not systematic loss [@Dornelas2014]. However, our comparison with replicated surveys from the same year warn that an important portion of the observed turnover of species might be due to pseudo-turnover (i.e. species difference between two surveys that were conducted by two surveyors during the same year on the same site). Given that the BDM program had invested significantly in developing reproducible methods and has continuously invested in quality control [@Plattner2004], the recorded pseudo-turnover was quite high. A potential explanation is that the replicated surveys are not conducted during the same days and in few cases the situation might have changed profoundly for example because the meadow was cut between the surveys. Furthermore, one of the botanist may not have identified difficult to identify species on species-level (i.e. differences in botanical skills), and--because we analysed only records that were identified on species level--it increased pseudo-turnover even tough both botanists detected the species. This must be taken into account when evaluating the presented results. For example, we find that sites with high species richness have higher turnover than sites with a low species richness. This seems biologically plausible, since the average species coverage in species-rich sites must be lower than in species-poor sites and species with low coverage probably have a higher turnover. At the same time, however, the result could also be simply due to pseudo-turnover, since species identification in species-rich sites is probably more difficult than in species-poor sites. However, for the presented results that suggest differences in colonization or local survival in relation to the species indicator values we can hardly imagine how this could be caused by methodological issues. We are thus confident that most of the results presented in this study are largely unbiased.

@@

However, the results of Fig. \@ref(fig:figconsequences) also show that the results of spatial comparisons must be interpreted carefully. When we applied inferred the spatial variation in oligotrophic species richness with the same covariates as in Roth et al. (2017) for different study periods, to infer how oligotrophic species richness is decreasing along the N deposition gradient, the relationship appeared to vary between study periods. The decrease gradually became steeper (more negative) over time, except for the first two study periods. Our first interpretation was that the N deposition effect became stronger over time. This was against our prediction that the effect of N deposition should become weaker over time, since N deposition is decreasing during the study period. Then we realized that species turnover was highest at low N deposition sites (Tab. \@ref(tab:turnovertab)). At low N deposition rates, colonizing species have below average indicator values for nutrients. It seems that the decrease in N deposition resulted in oligotrophic species replacing eutrophic species particularly at sites with low N deposition. This seems to explain why the decline in oligotrophic species richness inferred from spatial patterns of species richness and N deposition is becoming steeper over time. And this may be interpreted as evidence that plant communities are recovering at least at low deposition sites, and that the negative N deposition effects have not become stronger over time.
